---
title: "Assignment 2: Who Do We Know the Least About?"
subtitle: "Measuring uncertainty in American Community Survey data"
author: "Evelyn Zhong"
date: today
format:
  html:
    toc: true
    code-fold: false
execute:
  warning: false
  message: false
---

## Setup

```{r}
library(tidyverse)
library(tidycensus)
```

---

## The question

Every number the American Community Survey publishes is an estimate drawn from a
sample, not a count. That means every number comes with a margin of error — a
statement about how wrong it might be.

Most people ignore it. In this assignment, you won't.


**In this analysis, I look at how many households that received SNAP assistance with income were below poverty levels in the past 12 months across counties in California, and ask which communities we measure most and least reliably.**

---

## Choosing what to measure
```{r}
my_state <- "CA"                # two-letter abbreviation, in quotes
my_variable <- "B22003_003"     # the variable code, in quotes
my_state
my_variable
## Pulling the data


county_data <- get_acs(
  geography = "county",
  variables = my_variable,
  state = my_state,
  year = 2023,
  survey = "acs5"
)
```

Let's look at what came back.

```{r}
dim(county_data)
head(county_data)
```


**California has 58 counties, and dim(county_data) returns 58 rows.**

---

## How big is the margin of error?

A margin of error of 3,000 means something very different for an estimate of 20,000
than for an estimate of 200,000. To compare across places, we need it as a
percentage.

```{r}
county_data <- county_data %>%
  mutate(moe_pct = moe / estimate * 100)
```


Now sort to see which counties are measured least reliably.

```{r}
county_data %>%
  arrange(desc(moe_pct)) %>%
  select(NAME, estimate, moe, moe_pct) %>%
  head(10)
```


**The least reliable county appears to be Sierra County, California with a moe_pct of 130.8%. The margin of error is larger than the estimate itself so the true number of poverty-level SNAP households is obscured.**

---

## Sorting counties into reliability categories

```{r}
county_data <- county_data %>%
  mutate(reliability = case_when(
    moe_pct < 5  ~ "High confidence",
    moe_pct < 10 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))

count(county_data, reliability)
```
```{r}
summary(county_data$moe_pct)

county_data <- county_data %>%
  mutate(reliability = case_when(
    moe_pct < 10 ~ "High confidence",
    moe_pct < 25 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))
  
count(county_data, reliability)
```

**I set the cutoffs at 10% and 25% instead of the default 5% and 10%, based on the quantiles. The original thresholds pushed 41 of 58 counties into "Low confidence", but these new cutoffs roughly split counties into thirds. I chose 10 and 25 because it was closer to the actual quartiles and I played around with the numbers to test it out. It's worth noting that the distribution is right-skewed with a small number of very small counties like Sierra County pulling the average up.**

---

## Is the margin of error bigger in small places?

To answer this we need population and our variable side by side, in the same row.
That means a **wide** pull.

```{r}
county_wide <- get_acs(
  geography = "county",
  variables = c(myvar = my_variable,
                pop   = "B01003_001"),
  state = my_state,
  year = 2023,
  survey = "acs5",
  output = "wide"
)

head(county_wide)
```


```{r}
county_wide <- county_wide %>%
  mutate(moe_pct = myvarM / myvarE * 100) %>%
  mutate(reliability = case_when(
    moe_pct < 5  ~ "High confidence",
    moe_pct < 15 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))
```

Now compare the average population of counties in each reliability category.

```{r}
county_wide %>%
  group_by(reliability) %>%
  summarize(
    n_counties = n(),
    avg_population = mean(popE)
  )
```


**High confidence counties have an average population of around 5.2 million, compared to roughly 94,000 for lower confidence counties. Larger populations have larger ACS samples which help to produce smaller relative margins of error. Reliable SNAP-poverty estimates are likely only available for three of California 's largest counties. **

---

## What this means for policy

**The communities in California that we know least about include: Sierra, Mono, Alpine, Trinity, Plumas, Modoc, Glenn, Amador, Colusa, and San Benito. Sierra has a margin of error percentage of 130.8%, indicating that the estimate of 13 households has a margin of error larger than the number itself.**

**The communities of Mono and Alpine are also in the lower confidence bracket, likely due to the relatively lower average population. Communities with a higher average population place higher in confidence levels. The ACS is an ongoing sample survey across five years to build the estimates that we examine. Only a select few households are sampled for a given area, but the number of households with data increases as the population count rises. Los Angeles County will have more samples when compared to Sierra County, because the latter will have only a handful of survey respondents that even fall into the subpopulation of households with SNAP and below the poverty level.**

**If every county's number were to be treated as equally certain during funding allocation, smaller counties are likely to feel greater harm because these estimates are sensitive to sampling noise. Year to year, the numbers for a large population county like Los Angeles will budge slightly because the sample is large enough to average out the noise. However, a small county's number is more likely to swing which could affect their funding despite need still being great. **
